Type-Safe Modular Parsing

### 2016-12-29 paper/sections/Inheritance.tex
Added S3, inheritance-based approach.

\section{An Overview of \name}\label{sec:overview}

This section gives an overview of our library \name, motivates
additional problems that occur when parsing modular ASTs.  The first
problem is that traditional parsers using parser combinators use
hard-coded recursive calls. This problem is solved using
delegation-based techniques, which enable parsing OO ASTs.  The second
problem is how to achieve dual extensibility of ASTs (that is: how to
allow for both new language constructs and new operations over the
ASTs). This problem is solved using Object Algebras.

@@

\subsection{The Parsing Expression Problem}\label{subsec:overview-problem}
%\huang{done, added the attempt/failure/reason of extending conventional parsers}\bruno{You are not motivating the problem! You are going straight to
%  the solution without pointing out what the problem is first. What
%  you need to do is: First show what happens with conventional
%  parsers: at some point, if you add extensions the recursive calls
%  will be wrong. Then you show (in the next section) the solution:
% use delegation/open recursion.}

@@

In the following text we are going to illustrate the problems that
motivate our work via a minimalistic example inspired by Wadler's
Expression Problem~\cite{}.

\paragraph{Parsing a simple Arithmetic Language}
Suppose we want to parse a simple language with literals and
additions. The concrete syntax is as follows:
\bruno{Here's a good opportunity to both improve the paper and make it
more compact! This is the example that should be illustrated and
discussed at the end of Section 2. Then we can just create a figure to
refer to the code in Section 2 and refer again to the figure here. No
need for two different examples.}

\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int>
    \alt <expr> `+' <expr>
\end{grammar}
%\alt `(' <expr> `)'

\noindent It is straightforward to model ASTs by inheritance and write corresponding parsers for all cases:
%\huang{This is the first time to demonstrate code in this
%paper. Readers may not know what those combinators/types are. Refer
%to the library docs here? move the 'scala parsing library' section to
%the beginning as 'background'?}

\lstinputlisting[linerange=8-24]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_SIMPLE_EXPR

\noindent Note that we use \lstinline{Parser} as a type synonym for
\lstinline{PackratParser} in the rest of paper. 
The parser \inlinecode{pExpr} is able to parse 
expressions like \lstinline{"1+2"} correctly.

\paragraph{Dimensions of Extensibility}
At this point, one would like such code to be extensible in two dimensions, for example:
\begin{itemize}
\item \textbf{Extension 1}: adding a new case (or rather, a new construct) to \lstinline{Expr};
\item \textbf{Extension 2}: adding a new operation to \lstinline{Expr}.
\end{itemize}

@@

\lstinputlisting[linerange=28-30]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_SIMPLE_LAM

Unfortunatelly, extending the parser for the AST is not as easy.
Since we already have the parser for literals and additions, we would
like to build the new parser by reusing the old one. So we compose
\inlinecode{pVar}, which parses variables, with the old
parser. \inlinecode{pExtExpr} is the new function to parse
expressions.

\lstinputlisting[linerange=34-37]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_SIMPLE_EXT

\noindent This extended parser does not work properly to parse all possible expressions defined by the new grammar. For example, it fails to parse \inlinecode{"1 + x"} which is obviously valid.

\paragraph{Problem: Hard-coded Recursive Calls} 
Typical parsers defined by parser combinators follow the grammar
structure, they parse an input to build a corresponding AST. If the grammar is extended
with a new alternative, the corresponding parser has to be changed as
well. Thus the traditional style of developing parsers is not
extensible, and it is at odds with object-oriented programming ASTs,
which support the easy addition of new alternative classes.
The critical problem
of extending parsers directly is that the recursive calls in old
parsers are hard-coded. Specifically, \inlinecode{pAdd} makes two
recursive calls to parse two sub-expressions around
\lstinline{"+"}. Before the extension, we implemented it by using
\inlinecode{pExpr} for those recursive calls. It worked fine at that
time because \inlinecode{pExpr} covers all cases of the old
grammar. However, after the extension of variables, the proper
function for such recursive calls should be \inlinecode{pExtExpr}
instead of \inlinecode{pExpr}, because the new function
\inlinecode{pExtExpr} includes the new extension.

That is the reason why the extended parser fails to parse some valid
expressions such as \inlinecode{"1 + x"}. To resolve this issue,
\inlinecode{pAdd} needs to be aware of the change of grammar. There
are some possible solutions:

\begin{itemize}
    \item Keep the old code, rewrite the non-extensible function \inlinecode{pAdd}. \\Separate compilation is obtained but code reuse is lost, because every function may contain such recursive calls.
    \item Modify the old code, correct the recursive calls by proper functions. \\We can avoid duplication for code reuse, but separate compilation is sacrificed because of modification on the existing code.
    \item ??? \huang{third one? use dynamic languages?}
\end{itemize}

Because of the trade-off, neither of them are satisfying. The challenge of building semantically modular parsers requires fundamental change of the parser structure. We will introduce our solution in the next sections.

\subsection{Delegation-Based Parsing}\label{subsec:overview-delegation}

Hard-coded recursive calls prevent parsers from extensions.
This problem can be solved using delegation, which can be encoded with
\textit{open recursion}. The idea is simple: instead of making direct
recursive calls, an additional argument, that abstracts over the
recursive call, is passed to the parsers. This technique enables the
development of parsers that support OO-style extensibility of ASTs.

\paragraph{Open Parsing} Consider the expression language of only literals and additions in the previous section. The parser can be rewritten as follows:

\lstinputlisting[linerange=58-68]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_OPEN_BASE

According to the definition of type synonym \inlinecode{Open[T]},
every parsing function takes a parser of \inlinecode{Expr} as a
parameter, namely \inlinecode{self}. This parameter \inlinecode{self}
represents \emph{the parser} of expressions, which is dynamiclly
explained by the argument. Using such an abstract ``self-reference'' provides
extensibility to parsers, because the recursive calls are not
hard-coded here --- they are dynamic and can be changed corresponding
to the grammar!
We take \inlinecode{pAdd} as an example again. It uses
\inlinecode{self} instead of hard-coded \inlinecode{pExpr} for
recursive parsing. Depending on the real argument, \inlinecode{self}
can be a parser which covers only literals and additions, or a
parser which also supports variables.

\paragraph{Client Code} Since the type of parsing functions is
changed, we can not apply them directly. A \inlinecode{fix} function, which is the
the standard \textit{fixpoint combinator}, is used to `close' the open
recursion. After that, it can be used as an ordinary parser.
The client code below demonstrates the usage of new parsers.

%\huang{done}\bruno{Too much detail is being given here about open
%  recursion. In Section 2 we should talk only about \emph{how} to use
%open recursion, and not how open recursion is implemented. We
%can summarize, for example the API related to open recursion.
%Section 4 is where the implementation and detailed explanation
%about open recursion should come.
%}

\lstinputlisting[linerange=73-77]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_OPEN_USE

%\huang{done}\bruno{Please polish up code. You probably want to consistently use
%the variable name ``self'' whenever you want to refer to the whole
%parser (instead of ``e'' that you use at the moment). Also you should
%use ``Open'' (which you currently call ``Fix'') instead of using the
%longwinded ``(=> PackratParser[Expr]) => PackratParser[Expr]''}
%
%\huang{done}\bruno{Also, for consistency, why isn't ``pVar'' open? I.e. using a
%  self reference as well? I know that it is not recursive, but giving
%  it a different treatment makes it more confusing for users.}
%
%\huang{done}\bruno{You should use a more OO approach in the example.
%Rather than case classes show an Expr trait with pretty printing.
%This will enable you to talk about the extensibility and object
%algebras better later on.
%}

\paragraph{Extensibility} Similarly, we extend the grammar by adding variables as a new case. The new parser also extends the old one by adding \inlinecode{pVar} on it. Notice that the old code is not touched, hence separate compilation is obtained.

\lstinputlisting[linerange=82-85]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_OPEN_EXT

\haoyuan{Code without combinators?}

All recursive calls of parsing `an expression' are realized via the explicit parameter \inlinecode{self}. Once the \inlinecode{fix} function is applied, these calls will be updated appropriately. Parsing functions such as \inlinecode{pAdd} do not need to be rewritten. As a consequence, this parser recognizes all valid expressions of the new grammar, including \inlinecode{"1 + x"} which cannot be parsed previously.

\lstinputlisting[linerange=91-91]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_OPEN_EXT_USE

As we demonstrated, delegation encoded by open recursion is the key technique to obtain semantic modularity. It enables type-safe code reuse and separate compilation in our parsers. We will discuss this topic further in Section \ref{sec:openandparsing}.

@@

\lstinputlisting[linerange=111-111]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_ATTEMPT_EXPRWITHEVAL
But then the programmer has to define new classes for literals and additions that extend old ones and \lstinline{EvalExpr}. Specifically,
the critical point is that existing parsing code only generates \lstinline{Expr} objects with pretty printing, so modification on parsing is again required.

@@

\bruno{Put the following text somewhere?}
\paragraph{\name} Our library is constructed by only around twenty lines of code. We have wrapped up the type of parsers
to make them modular and extensible, and we use our custom combinator \lstinline{<|>} for alternative composition of those
modular ``open parsers'' with longest match, built on \lstinline{|||}. Table~\ref{tab:parser} shows the API that \name provides for modular parsing.

>>>>

\section{Modularizing Parsers with Inheritance}\label{sec:inheritance}

This section introduces the problem that motivates our work. In Figure~\ref{fig:packrat-arith} we have observed
that Packrat parsers can be defined in a convenient way, with recursive calls. The example is actually inspired
by Walder's Expression Problem. Similarly, our goal is to build parsers that are extensible in some dimensions.

\subsection{Dimensions of Extensibility}
We continue with the expression language of literals and additions. At this point,
one would like to have two kinds of extensions:

%\section{An Overview of \name}\label{sec:overview}
%
%This section gives an overview of our library \name, motivates
%additional problems that occur when parsing modular ASTs.  The first
%problem is that traditional parsers using parser combinators use
%hard-coded recursive calls. This problem is solved using
%delegation-based techniques, which enable parsing OO ASTs.  The second
%problem is how to achieve dual extensibility of ASTs (that is: how to
%allow for both new language constructs and new operations over the
%ASTs). This problem is solved using Object Algebras.

@@

%\huang{done, added the attempt/failure/reason of extending conventional parsers}\bruno{You are not motivating the problem! You are going straight to
%  the solution without pointing out what the problem is first. What
%  you need to do is: First show what happens with conventional
%  parsers: at some point, if you add extensions the recursive calls
%  will be wrong. Then you show (in the next section) the solution:
% use delegation/open recursion.}

@@

\begin{itemize}
\item \textbf{Extension 1}: adding a new case (or rather, a new construct) to \lstinline{Expr};
\item \textbf{Extension 2}: adding a new operation to \lstinline{Expr}.
\end{itemize}

@@

\lstinputlisting[linerange=95-97]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_SIMPLE_LAM

And since we already have the parser for literals and additions, we would
like to build the new parser by reusing the old one. Here one may quickly come up
with the following attempt, where a new parser is defined for \lstinline{Var}, then composed
with \lstinline{pExpr}:

\lstinputlisting[linerange=62-63]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=BAD_ATTEMPT

Unfortunately, it fails to parse some expressions like \lstinline{"1 + x"}, which is obviously valid in the new grammar.
The reason is that \lstinline{pAdd} makes two recursive calls to parse sub-expressions, by using \lstinline{pExpr}, which
covers both cases in the old grammar. But the newly added case \lstinline{pVar} is not observed by the recursive \lstinline{pExpr},
hence the parser does not work as expected. Still, one may continue to change the recursive call in \lstinline{pAdd} with \lstinline{pExtExpr},
but modification on existing code sacrifices separate compilation, as mentioned by the Expression Problem.

\subsection{Inheritance-Based Approach}\label{subsec:inheritance-approach}

Fortunately in Scala we can make use of inheritance for modular parsing. Specifically,
the new parser can be defined in an enclosing trait that extends
the old one:

\lstinputlisting[linerange=68-73]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=INHERITANCE_APPROACH
Now \lstinline{ExtParser} successfully represents the parser for the extended language, because Scala uses dynamic dispatch for
method overriding in inheritance. When the input \lstinline{"1 + x"} is fed to the parser \lstinline{this.pExpr}, it firstly delegates
the work to \lstinline{super.pExpr}, which parses literals and additions. However, the recursive call \lstinline{pExpr} in \lstinline{pAdd}
actually refers to \lstinline{this.pExpr} again, due to dynamic dispatch. It implies that the recursive call has included the extension. This
elegant inheritance-based approach opens the recursion of parsers in a modular way.

Notice that Scala has a linearized-style multiple inheritance for traits. It can be very helpful when composing several languages. Suppose now
we want to compose the parsers for expressions from pre-defined languages \lstinline{LanguageA} and \lstinline{LanguageB} using alternative.
The use of keyword \lstinline{super} is able to
specify the implementation from inheritance, like:

\lstinputlisting[linerange=78-84]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=MULTIPLE_INHERITANCE

As we demonstrated, inheritance is the key technique to obtain semantic modularity.
It enables type-safe code reuse and separate compilation for parsing OO style ASTs with extension 1.

@@

\lstinputlisting[linerange=178-178]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_ATTEMPT_EXPRWITHEVAL
But then the programmer has to define new classes for literals and additions that extend old ones and \lstinline{EvalExpr}. Specifically,
the critical point is that existing parsing code only generates \lstinline{Expr} objects with pretty printing, so modification on parsing is again required.

@@

### 2016-12-29 paper/sections/OpenRecursion.tex
Added S3, inheritance-based approach.

\lstinputlisting[linerange=31-37]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=OPENRECURSION_FIB

@@

\lstinputlisting[linerange=41-47]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=OPENRECURSION_FIB2
\bruno{Fix should be replaces by Open. Please coordinate with Huang to
ensure that your code and his code are coherent (i.e you use same
names and same conventions),}
\noindent This alternative implementation has an extra argument
\lstinline{self}, which serves as an explicit self-reference. In the
definition of the function, direct recursive calls are replaced by
calls to the self reference.

@@

\lstinputlisting[linerange=51-51]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=OPENRECURSION_FIB_FIX

@@

\lstinputlisting[linerange=27-27]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=OPENRECURSION_FIX
\bruno{can be defined in a single line.}
\noindent Note that the function \lstinline{f} has a by-name parameter
of type \lstinline{"=>} \lstinline{T"}. This is necessary to ensure
termination of well-founded recursive calls. In such cases,
there are some base cases (like the first two in
\lstinline{fib2}) which stop the recursion in their branches, and the
evaluation is finally reduced to those cases.

@@

\bruno{We may want to use a better example than what is next.}
Suppose now we want to make use of open recursion for additional operations. We first define a combinator which combines the results of two generators in a pair:
\lstinputlisting[linerange=55-58]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=OPENRECURSION_COMPOSE

Using Scala implicit class, \lstinline{*} can be used as an infix operator for composition in convenience.
Then we define a generator which prints out a number as a string:
\lstinputlisting[linerange=62-62]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=OPENRECURSION_SHOW

@@

Next we define a more general combinator \lstinline{merge} for open recursion. It is similar to the \lstinline{*} above, but instead of putting two values in a pair by default, \lstinline{merge} takes a parameter called \lstinline{op}, which tells the associative operation of two results. The definition is straightforward:
\lstinputlisting[linerange=66-67]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=OPENRECURSION_MERGE

>>>>

\lstinputlisting[linerange=24-30]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=OPENRECURSION_FIB

@@

\lstinputlisting[linerange=34-40]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=OPENRECURSION_FIB2
\bruno{Fix should be replaces by Open. Please coordinate with Huang to
ensure that your code and his code are coherent (i.e you use same
names and same conventions),}
\noindent This alternative implementation has an extra argument
\lstinline{self}, which serves as an explicit self-reference. In the
definition of the function, direct recursive calls are replaced by
calls to the self reference.

@@

\lstinputlisting[linerange=44-44]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=OPENRECURSION_FIB_FIX

@@

\lstinputlisting[linerange=20-20]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=OPENRECURSION_FIX
\bruno{can be defined in a single line.}
\noindent Note that the function \lstinline{f} has a by-name parameter
of type \lstinline{"=>} \lstinline{T"}. This is necessary to ensure
termination of well-founded recursive calls. In such cases,
there are some base cases (like the first two in
\lstinline{fib2}) which stop the recursion in their branches, and the
evaluation is finally reduced to those cases.

@@

\bruno{We may want to use a better example than what is next.}
Suppose now we want to make use of open recursion for additional operations. We first define a combinator which combines the results of two generators in a pair:
\lstinputlisting[linerange=48-51]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=OPENRECURSION_COMPOSE

Using Scala implicit class, \lstinline{*} can be used as an infix operator for composition in convenience.
Then we define a generator which prints out a number as a string:
\lstinputlisting[linerange=55-55]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=OPENRECURSION_SHOW

@@

Next we define a more general combinator \lstinline{merge} for open recursion. It is similar to the \lstinline{*} above, but instead of putting two values in a pair by default, \lstinline{merge} takes a parameter called \lstinline{op}, which tells the associative operation of two results. The definition is straightforward:
\lstinputlisting[linerange=59-60]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=OPENRECURSION_MERGE

### 2016-12-29 paper/sections/PackratParsing.tex
Added S3, inheritance-based approach.

This section discusses the algorithmic challenges introduced by modular parsing and argues that parser combinators based on Packrat Parsing 
are suitable to address those challenges.

@@

\subsection{Packrat Parsing}
Fortunately some more advanced parsing techniques such as Packrat parsing~\cite{Ford2002}, address the limitations of simple parser combinators such as Parsec. Packrat parsers use memoization to record the result of applying each parser at each position of the input, so that repeated computation is eliminated.  Moreover, theoretically the algorithm behind Packrat parsers supports both direct and indirect left-recursion~\cite{warth2008}. All of these properties are very suitable for modularity, thus we decided to use Packrat parsers as the underlying parsing technique in our modular parsing library.
\begin{comment}
It is worth mentioning that the choice of parser combinators will not
affect the other parts of our library. One can choose other parser
combinators like Parsec, in cases that the performance and supporting
of left-recursion are not major concerns. A different library can even build a new
\name with fancy features or higher efficiency.
\end{comment}
In the current version of Scala (2.11), Packrat Parsing is included
in the Scala standard parser combinator library, which is an external jar file. The library provides a number of
parser combinators, together with some basic parsers. Below we present an example to illustrate Scala parsers.

\bruno{So, we make a big fuss identifying the problems with other 
parser combinators in the previous subsection and now we present 
an example that does not suffer from any of the previous problems!
What makes sense here is an example that uses: a \emph{left-recursive grammar}; the \emph{longest match combinator}; and where we can argue that 
there's no need for the user to do manual backtracking. Just use 
the example with the left-recursive grammar for parsing integers and additions and argue that none of the limitations 
discussed previously applies!
}

\paragraph{An Example}
A Packrat parser has type \lstinline{PackratParser[E]} for some
\lstinline{E}, which indicates the type of results it produces.
%\bruno{the representation of what?}
And we use \lstinline{lexical} for its lexing.

\lstinputlisting[linerange=8-12]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=PACKRAT_EXAMPLE
\noindent
Note that the code should be defined in an enclosing type that extends
\lstinline{scala.util.parsing.combinator.syntactical.StandardTokenParsers}
for lexing and
\lstinline{scala.util.parsing.combinator.PackratParsers}.
\bruno{Well, if we need to present the lexing code, then it is better 
to present the code for the enclosing type as well. Instead of 
just saying this in the text.}
The first two lines are used for lexing. The third line defines a
parser which can, for instance, parse \lstinline{"str(5)"} and produce
the integer \lstinline{5}, where \lstinline{numericLit} is just a basic parser
in the API. Table~\ref{tab:packrat} gives a closer look at part of the API that 
we have used in our work.

@@

\begin{table}[t]
\begin{tabular}{l}
\hline
\begin{lstlisting}
type OpenParser[Alg, R, E] = Alg => (=> R) => PackratParser[E]
\end{lstlisting} \\
\hspace{.2in}- The type constructor for modular parsers. \\
\hline
\begin{lstlisting}
def parse(p: Parser[_]): String => Unit
\end{lstlisting} \\
\hspace{.2in}- A function that applies a parser to an input, and prints the result as a string. \\
\hline
\begin{lstlisting}
implicit def <|>(x: OpenParser[Alg1, R, E], y: OpenParser[Alg2, R, E])
    : OpenParser[Alg1 with Alg2, R, E]
\end{lstlisting} \\
\hspace{.2in}- Open parser combinator for alternative with longest match composition. \\
\hline \\
\end{tabular}
\caption{Our parser combinator library \name API.\haoyuan{I don't know if I can present <|> like this.}}\label{tab:parser}
\end{table}

@@

\lstinputlisting[linerange=16-19]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=PACKRAT_RUNPARSER
Now \lstinline{parse(p)("str(5)")} produces \lstinline{5} as expected.

>>>>

This section discusses the algorithmic challenges introduced by modular parsing and argues that parser combinators from Packrat Parsing
are suitable to address those challenges.

@@

\subsection{Packrat Parsing}
Fortunately some more advanced parsing techniques such as Packrat parsing~\cite{Ford2002}, address the limitations of simple parser combinators such as Parsec. Packrat parsers use memoization to record the result of applying each parser at each position of the input, so that repeated computation is eliminated.  Moreover, theoretically the algorithm behind Packrat parsers supports both direct and indirect left-recursion~\cite{warth2008}. All of these properties are very suitable for modularity, thus we decided to use Packrat parsers as the underlying parsing technique in our modular parsing library.
\begin{comment}
It is worth mentioning that the choice of parser combinators will not
affect the other parts of our library. One can choose other parser
combinators like Parsec, in cases that the performance and supporting
of left-recursion are not major concerns. A different library can even build a new
\name with fancy features or higher efficiency.
\end{comment}
In the current version of Scala (2.11), Packrat Parsing is included
in the Scala standard parser combinator library. The library provides a number of
parser combinators, together with some basic parsers. Below we present an example to illustrate Scala parsers.
%
%\bruno{So, we make a big fuss identifying the problems with other
%parser combinators in the previous subsection and now we present
%an example that does not suffer from any of the previous problems!
%What makes sense here is an example that uses: a \emph{left-recursive grammar}; the \emph{longest match combinator}; and where we can argue that
%there's no need for the user to do manual backtracking. Just use
%the example with the left-recursive grammar for parsing integers and additions and argue that none of the limitations
%discussed previously applies!
%}

\paragraph{Parsing a simple Arithmetic Language}
Suppose we want to parse a simple language with literals and
additions. The concrete syntax is as follows:
%\bruno{Here's a good opportunity to both improve the paper and make it
%more compact! This is the example that should be illustrated and
%discussed at the end of Section 2. Then we can just create a figure to
%refer to the code in Section 2 and refer again to the figure here. No
%need for two different examples.}

\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int>
    \alt <expr> `+' <expr>
\end{grammar}
%\alt `(' <expr> `)'

It is straightforward to model ASTs by inheritance and write corresponding parsers for all cases.
Note that a parser has type \lstinline{PackratParser[E]} for some
\lstinline{E}, which indicates the type of results it produces. Figure~\ref{fig:packrat-arith} shows the abstract syntax and
the parser code that generates expressions as ASTs.

\begin{figure}[t]
\centering
\lstinputlisting[linerange=7-27]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_SIMPLE_EXPR
\caption{Packrat Parsing for an arithmetic language.}\label{fig:packrat-arith}
\end{figure}

In the trait \lstinline{BaseParser}, \lstinline{lexical} is used for lexing. \lstinline{pLit} can parse an integer for \lstinline{Lit},
whereas \lstinline{pAdd} handles addition and creates an object of \lstinline{Add}. It parses two sub-expressions by calling \lstinline{pExpr}
recursively. Finally \lstinline{pExpr} composes \lstinline{pLit} and \lstinline{pAdd} using the alternative combinator with longest match.
Table~\ref{tab:packrat} gives a closer look at part of the API that we have used in our work.

@@

@@

\lstinputlisting[linerange=52-57]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=PACKRAT_RUNPARSER

It is worth mentioning that such a grammar is left-recursive, yet well supported by Packrat parsers. It also employs automatic
longest match composition by using the library function \lstinline{|||}. Finally, Packrat parsers do not suffer from the backtracking
problem, as the technique uses memoization to record intermediate parsing results. We will use \lstinline{Parser} as a type synonym for \lstinline{PackratParser} in the rest of paper.
